Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-15 prop_letter.tex
prop_letter.tex \section{Gleason} Updating...

The quantum probability measure~$\mu_{1}$ maps some projectors to
$0$ or $0.5$, which are replaced by $\left[0,0\right]$ or $\left[0.5,0.5\right]$
to get $\bar{\mu}_{3}$. The way we replace two numbers by the smallest
intervals containing them can be generalized if we carefully pick
a set of intervals and the rule to select the correct interval when
intervals are overlapping. For example, consider the following collection
of intervals: 
\begin{eqnarray}
\mathscr{I}_{n} & = & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\{\imposs,\necess\}
\end{eqnarray}
which are overlapping intervals with length~$1/n$ except the intervals
at length~$\left(2n\right)^{-1}$ at each end. Since these intervals
are overlapping, we define the non-overlapping interior part of a
interval according to the following rule.\begin{subequations}
\begin{eqnarray}
\textrm{Int}\left(\imposs\right) & = & \imposs\,,\\
\textrm{Int}\left(\left[0,\tfrac{1}{2n}\right]\right) & = & \left(0,\tfrac{0.5}{2n}\right]\,,\\
\textrm{Int}\left(\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\right) & = & \left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]\nonumber \\
 &  & \textrm{for even }0<k<2n\,,\\
\textrm{Int}\left(\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\right) & = & \left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)\nonumber \\
 &  & \textrm{for odd }k\,,\\
\textrm{Int}\left(\left[\tfrac{2n-1}{2n},1\right]\right) & = & \left[\tfrac{2n-0.5}{2n},1\right)\,,\\
\textrm{Int}\left(\necess\right) & = & \necess\,.
\end{eqnarray}
\end{subequations}Hence, given a real-valued probability~$x$, we
will replace it with an interval~$\iota_{n}\left(x\right)=[\ell,r]$
if $x\in\textrm{Int}\left([\ell,r]\right)$.

After we have a rule to replace a real number by an interval, any
Born rule probability~$\mu_{\rho}^{\mathrm{B}}$ on a three-dimensional
Hilbert space can be extended to a QIVPM~$\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
where $\circ$ is the function concatenation. For example, when $n=2$,
we can summarize the set of intervals~$\mathscr{I}_{2}$ and their
interior in Table~\ref{tab:I2}.
\begin{table}
\caption{\label{tab:I2}Intervals in $\mathscr{I}_{2}$ and their interior.}
\begin{tabular}{cc}
\toprule 
$[\ell,r]$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline
\midrule
\midrule 
$\imposs$ & $\imposs$\tabularnewline
\midrule 
$\left[0,\tfrac{1}{4}\right]$ & $\left(0,\tfrac{1}{8}\right]$\tabularnewline
\midrule 
$\left[0,{\textstyle \tfrac{1}{2}}\right]$ & $\left(\tfrac{1}{8},\tfrac{3}{8}\right)$\tabularnewline
\midrule 
$\left[\tfrac{1}{4},\tfrac{3}{4}\right]$ & $\left[\tfrac{3}{8},\tfrac{5}{8}\right]$\tabularnewline
\midrule 
$\left[\tfrac{1}{2},1\right]$ & $\left(\tfrac{5}{8},\tfrac{7}{8}\right)$\tabularnewline
\midrule 
$\left[\tfrac{1}{4},1\right]$ & $\left[\tfrac{7}{8},1\right)$\tabularnewline
\midrule 
$\necess$ & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\end{table}
Given two quantum probability measure~$\mu_{1}$ and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$,
where $\frac{\mathbb{1}}{3}=\frac{1}{3}\left(\proj{0}+\proj{1}+\proj{2}\right)$,
it is then easy to compute $\iota_{2}\circ\mu_{1}$ and $\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$
on the minimal subspace of events containing $\proj{0}$, $\proj{1}$,
and $\proj{2}$, $\events_{C}$ as in Table~\ref{tab:Compute-the-QIVPM}.
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\iota_{2}\circ\mu_{1}$ and
$\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$ based on $\mu_{1}$
and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$ on $\events_{C}$.}
\begin{tabular}{ccccc}
\toprule 
$P$ & $\mu_{1}\left(P\right)$ & $\iota_{2}\left(\mu_{1}\left(P\right)\right)$ & $\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline
\midrule
\midrule 
$\mathbb{0}$ & $0$ & $\imposs$ & $0$ & $\imposs$\tabularnewline
\midrule 
$\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{2}$ & $0$ & $\imposs$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{2}$ & $1$ & $\necess$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}$ & $1$ & $\necess$ & $1$ & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\end{table}
Verifying $\iota_{2}\circ\mu_{1}$ and $\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$
are QIVPMs are also straightforward.

Given any state~$\rho$ satisfying $0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$
for any $P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $,
called totally mixed state, we know not only $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is a QIVPM, but also the collection of states consistent it get closer
and closer to $\rho$ as the intervals get narrower and narrower,
eventually collapsing to $\rho$ when the intervals become infinitely
precise real numbers. To state this result, we need to define a notion
of distance based on the following norm of a matrix~$A$ as $\left\Vert A\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$~\citep{544199}.
If a state~$\rho'$ is consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
its distance with $\rho$, $\left\Vert \rho-\rho'\right\Vert $, must
be bounded by $\left(12n\right)^{-1}$. Reversely, if the distance
between $\rho$ and $\rho'$ is bounded by $1/n$, then $\rho'$ is
consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$. In other
words, for the states consistent with this QIVPM, $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho'$ as $n$ goes to infinity.

For example, since $\mu_{1}\left(\proj{2}\right)=0$, any state corresponds
to $\mu_{1}$ is not totally mixed state. On the other hand, $\mathbb{1}/3$
is a totally mixed state. \yutsung{To be continue...}

@@

\noindent However, if the intervals are only with finite-precision,
there are some QIVPMs which cannot be consistent with any state. Given
a three-dimensional Hilbert space $\Hilb$ with an orthonormal basis
$\left\{ \ket{0},\ket{1},\ket{2}\right\} $, let $\mathscr{I}_{1}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$,
and \begin{subequations} 
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}(\mathbb{0})=\bar{\mu}(\proj{0})\\
 & =\bar{\mu}(\proj{\ps})=\bar{\mu}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}(\mathbb{1})=\bar{\mu}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}(\mathbb{1}-\proj{\ps})=\bar{\mu}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations} It is straightforward to verify that $\bar{\mu}$
is a QIVPM. Then we will prove $\coreBorn\left(\bar{\mu},\events\right)$
is an empty set by contradiction. Suppose there is a state $\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j}>0$. Since $\mu_{\rho}^{\mathrm{B}}(P)\in\bar{\mu}(P)$
for any~$P$, the state~$\rho$ must satisfy: 
\begin{equation}
\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps'})=0
\end{equation}
which implies $\ip{\phi_{j}}{0}=\ip{\phi_{j}}{\ps}=\ip{\phi_{j}}{\ps'}=0$
for all $j$. However, the last equation never holds. Thus, there
is no state consistent with the QIVPM on all projectors~$P\in\events$,
so Gleason's theorem cannot be extended to general QIVPMs because
of this example.

>>>>

The quantum probability measure~$\mu_{1}$ maps some projectors to
$0$ or $0.5$, which are replaced by $\left[0,0\right]$ or $\left[0.5,0.5\right]$
to get $\bar{\mu}_{3}$. The way we replace two numbers by the smallest
intervals containing them can be generalized if we carefully pick
a set of intervals and the rule to select the correct interval when
intervals are overlapping. For example, consider the following collection
of intervals: 
\begin{eqnarray}
\mathscr{I}_{n} & = & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\{\imposs,\necess\}
\end{eqnarray}
which are overlapping intervals with length~$1/n$ except the intervals
with length~$\left(2n\right)^{-1}$ at each end. Since these intervals~$[\ell,r]$
are overlapping, we define the non-overlapping interior part of a
interval~$\textrm{Int}\left([\ell,r]\right)$ as in Table~\ref{tab:I2}
which also contains an concrete example for $\mathscr{I}_{2}$.
\begin{table}
\caption{\label{tab:I2}Intervals in $\mathscr{I}_{n}$ and their interior
with an concrete example for $\mathscr{I}_{2}$. Although intervals~$[\ell,r]$
are overlapping, their interior~$\textrm{Int}\left([\ell,r]\right)$
are non-overlapping.}
\begin{tabular}{cccc}
\toprule 
\multicolumn{2}{c}{$\mathscr{I}_{n}$} & \multicolumn{2}{c}{$\mathscr{I}_{2}$}\tabularnewline
\midrule 
$[\ell,r]$ & $\textrm{Int}\left([\ell,r]\right)$ & $[\ell,r]$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline
\midrule
\midrule 
$\imposs$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline
\midrule 
$\left[0,\tfrac{1}{2n}\right]$ & $\left(0,\tfrac{0.5}{2n}\right]$ & $\left[0,\tfrac{1}{4}\right]$ & $\left(0,\tfrac{1}{8}\right]$\tabularnewline
\midrule 
$\begin{gathered}\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\\
\textrm{for odd }k
\end{gathered}
$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$ & $\left[0,{\textstyle \tfrac{1}{2}}\right]$ & $\left(\tfrac{1}{8},\tfrac{3}{8}\right)$\tabularnewline
\midrule 
\multirow{2}{*}{$\begin{gathered}\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\\
\textrm{for even }k\\
0<k<2n
\end{gathered}
$} & \multirow{2}{*}{$\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$} & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$ & $\left[\tfrac{3}{8},\tfrac{5}{8}\right]$\tabularnewline
\cmidrule{3-4} 
 &  & $\left[\tfrac{1}{2},1\right]$ & $\left(\tfrac{5}{8},\tfrac{7}{8}\right)$\tabularnewline
\midrule 
$\left[\tfrac{2n-1}{2n},1\right]$ & $\left[\tfrac{2n-0.5}{2n},1\right)$ & $\left[\tfrac{1}{4},1\right]$ & $\left[\tfrac{7}{8},1\right)$\tabularnewline
\midrule 
$\necess$ & $\necess$ & $\necess$ & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\end{table}
Hence, given a real-valued probability~$x$, we could easily replace
it with an interval~$\iota_{n}\left(x\right)=[\ell,r]$ if $x\in\textrm{Int}\left([\ell,r]\right)$.

After we have a rule to replace a real number by an interval, any
Born rule probability~$\mu_{\rho}^{\mathrm{B}}$ on a three-dimensional
Hilbert space can be extended to a QIVPM~$\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
where $\circ$ is the function concatenation. For example, when $n=2$,
two quantum probability measure~$\mu_{1}$ and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$
can induce two QIVPMs, where $\mathbb{1}/3=\left(\proj{0}+\proj{1}+\proj{2}\right)/3$.
The interval-valued probability of the induced QIVPMs, $\iota_{2}\left(\mu_{1}\left(P\right)\right)$
and $\iota_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$,
for projector $P$ in the minimal subspace of events containing $\proj{0}$,
$\proj{1}$, and $\proj{2}$, $\events_{C}$, are computed straightforwardly
in Table~\ref{tab:Compute-the-QIVPM}.
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\iota_{2}\circ\mu_{1}$ and
$\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$ based on $\mu_{1}$
and $\mu_{\mathbb{1}/3}^{\mathrm{B}}$ on $\events_{C}$.}
\begin{tabular}{ccccc}
\toprule 
$P$ & $\mu_{1}\left(P\right)$ & $\iota_{2}\left(\mu_{1}\left(P\right)\right)$ & $\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline
\midrule
\midrule 
$\mathbb{0}$ & $0$ & $\imposs$ & $0$ & $\imposs$\tabularnewline
\midrule 
$\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\proj{2}$ & $0$ & $\imposs$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{0}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{1}$ & $0.5$ & $\left[\frac{1}{4},\frac{3}{4}\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}-\proj{2}$ & $1$ & $\necess$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline
\midrule 
$\mathbb{1}$ & $1$ & $\necess$ & $1$ & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\end{table}
Verifying $\iota_{2}\circ\mu_{1}$ and $\iota_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}}$
are QIVPMs are also straightforward.

A state~$\rho$ is called totally mixed state if it satisfies $0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$
for any $P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $.
Given any totally mixed state~$\rho$, we know not only $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is a QIVPM, but also the collection of states consistent it get closer
and closer to $\rho$ as the intervals get narrower and narrower,
eventually collapsing to $\rho$ when the intervals become infinitely
precise real numbers. To state this result, we need to define a notion
of distance based on the following norm of a matrix~$A$ as $\left\Vert A\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$~\citep{544199}.
If a state~$\rho'$ is consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$,
its distance with $\rho$, $\left\Vert \rho-\rho'\right\Vert $, must
be bounded by $1/n$. Reversely, if the distance between $\rho$ and
$\rho'$ is bounded by $\left(12n\right)^{-1}$, then $\rho'$ is
consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$. For example,
since $\mu_{1}\left(\proj{2}\right)=0$, any state corresponds to
$\mu_{1}$ is not totally mixed state. On the other hand, $\mathbb{1}/3$
is a totally mixed state. Consider $\rho'=\left(\frac{1}{3}+\frac{1}{24}\right)\proj{0}+\left(\frac{1}{3}-\frac{1}{24}\right)\proj{1}+\frac{1}{3}\cdot\proj{2}$.
Since $\left\Vert \rho-\rho'\right\Vert \le\frac{1}{12\cdot2}$, we
should have $\rho'\in\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}},\events\right)$.
Although verifying the general case is non-trivial, it is easy to
see $\mu_{\rho'}^{\mathrm{B}}\left(P\right)\in\iota_{2}\left(\mu_{\mathbb{1}/3}^{\mathrm{B}}\left(P\right)\right)$
for $P\in\events_{C}$. As $n$ goes to infinity, only the states
very close to $\rho$ is guarantied belong to $\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}},\events\right)$,
and the states consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho$.

@@

\noindent However, if the intervals are only with finite-precision,
there are some QIVPMs which cannot be consistent with any state. Given
a three-dimensional Hilbert space $\Hilb$ with an orthonormal basis
$\left\{ \ket{0},\ket{1},\ket{2}\right\} $, let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$,
and \begin{subequations} 
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}(\mathbb{0})=\bar{\mu}(\proj{0})\\
 & =\bar{\mu}(\proj{\ps})=\bar{\mu}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}(\mathbb{1})=\bar{\mu}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}(\mathbb{1}-\proj{\ps})=\bar{\mu}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations} It is straightforward to verify that $\bar{\mu}$
is a QIVPM. Then we will prove $\coreBorn\left(\bar{\mu},\events\right)$
is an empty set by contradiction. Suppose there is a state $\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j}>0$. Since $\mu_{\rho}^{\mathrm{B}}(P)\in\bar{\mu}(P)$
for any~$P$, the state~$\rho$ must satisfy: 
\begin{equation}
\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps'})=0
\end{equation}
which implies $\ip{\phi_{j}}{0}=\ip{\phi_{j}}{\ps}=\ip{\phi_{j}}{\ps'}=0$
for all $j$. However, the last equation never holds. Thus, there
is no state consistent with the QIVPM on all projectors~$P\in\events$,
so Gleason's theorem cannot be extended to general QIVPMs because
of this example.
